Cosmology with photometric weak lensing surveys: constraints with redshift tomography of convergence peaks and moments

### 2016-05-03 photoz_paper/paper.tex
last edits before submission

When the dimension $N_b$ of the summary statistics space is large, numerical issues can arise in the estimation of the parameter error bars if the covariance matrix $\bbh{C}$ is measured from simulations. When $N_r$ independent realizations are used to estimate $\bbh{C}$, its inverse $\bbh{\Psi}$ is biased by a constant factor (see \citep{Hartlap07,Taylor12,Taylor14}) which can be taken into account. When the bias correction is applied, we can calculate the expectation value of the covariance estimator (\ref{meth:parcovestimator}) (see again\citep{Taylor14})
\begin{equation}
\label{meth:covdegradation}
\langle\bbh{\Sigma}\rangle = \left(\frac{N_r-N_b+N_p-1}{N_r-N_b-2}\right)\bb{\Sigma}
\end{equation}   
%
where $\bb{\Sigma}$ is the asymptotic covariance one obtains with an infinite number of realizations and $N_p=3$ is the number of parameters we are estimating. The scatter of the parameter estimates (\ref{meth:parestimate}) on the other hand scales as \citep{Taylor14}
\begin{equation}
\label{meth:estdegradation}
\langle{\rm Cov}(\bbh{p})\rangle = \left(\frac{N_r-2}{N_r-N_b+N_p-2}\right)\bb{\Sigma}
\end{equation}
%
Although equations (\ref{meth:covdegradation}) and (\ref{meth:estdegradation}) agree in the limit $N_r\rightarrow\infty$, they can be different when a finite number of realizations is used. The degradation factor in the parameter covariance estimate in (\ref{meth:covdegradation}) is of order $1+(1+N_p)/N_r$, while the scatter of the estimates $\bbh{p}$ is of order $1+(N_b-N_p)/N_r$. These numbers can be very different if $N_b$ is large. This means that the parameter error bar estimate (\ref{meth:covdegradation}) is too conservative if $N_b/N_r$ is of order unity. This could be likely the case in our analysis because of the inclusion of tomography information. If we bin the single redshift summary statistic with $N_{\rm st}$ intervals, and consider $N_z$ redshift bins, this can lead to a summary statistic vector of size $N_b = O(N_{\rm st}N_z^2)$ for the power spectrum and $N_b = O(N_{\rm st}N_z)$ for the remaining statistics. This can become quickly comparable with $N_r=16000$ once more redshift bins or a finer binning of the summary statistic are considered. In order to avoid these error degradation issues, we apply dimensionality reduction techniques to the summary statistics we are considering. Even if these techniques might not play a vital role in this work, as the maximum $N_b/N_r$ ratio we use is of order 1\%, they will definitely be relevant in future experiments when using finely binned summary statistics or when combining different cosmological probes. We explain the dimensionality reduction techniques we adopted in the next paragraph.

@@

\subsection{Dimensionality reduction}
We apply a Principal Component Analysis (see \citep{astroMLText} for example) to reduce the dimensionality of our summary statistics while preserving the cosmological information content. The model statistic $\bb{d}(\bb{p})$ can be regarded as a $P\times N_b$ matrix $d_{pi}$. Consider the whitened model matrix
\begin{equation}
\label{meth:whitening}
\Delta_{pi} = P\frac{d_{pi}}{\sum_{q=1}^Pd_{qi}} - 1.
\end{equation} 
%
We perform a Singular Value Decomposition (SVD) of $\Delta$ 
\begin{equation}
\bb\Delta = \bb{L}\bb{\Lambda} \bb{R}
\end{equation}
%
where $L$ is $P\times Q$, $\bb{\Lambda}={\rm diag}(\Lambda_1,...,\Lambda_Q)$ and $\bb{R}$ is $Q\times N_b$ and $Q={\min}(P,N_b)$. $R_{ni}$ is the $i$-th component of the $n$-th basis vector in statistics space. The singular value $\Lambda_n$ is the variance of the whitened summary statistic along the $n$-th basis vector. We assume that only summary statistic projections on the first $N_c$ basis vectors contain relevant cosmological information, where $N_c<N_b$ is a number that has to be determined from the simulations. Let $\bb{R}(N_c)$ be a matrix made of the first $N_c$ rows of $\bb{R}$ (we assume that the singular values $\Lambda_i$ are sorted). We define the PCA projection of a summary statistic $\bbh{d}$ on $N_c$ principal components as 
\begin{equation}
\label{meth:pcaprojection}
\h{d}_n(N_c) = \sum_{i=1}^{N_b}(\bb{R}(N_c))_{ni}\left(P\frac{\h{d}_i}{\sum_{p=1}^P d_{pi}}-1\right) 
\end{equation} 
%
Through the above procedure, we hope to capture the cosmological information contained in $\bbh{d}$ by projecting it on the $N_c<N_b$ principal components that vary the most with cosmology parameters. In the next section we outline the main results of this work.

@@

\subsection{Priors from CMB experiments}
In this section we describe how we included prior knowledge of cosmological parameters from previous Cosmic Microwave Background (CMB) observations, such as Planck \citep{PlanckCosmo}. CMB experiments provide tight constraints on $(\Omega_m,\sigma_8)$, but they are not sensitive to dark energy parameters such as $w$. Nevertheless, prior knowledge of $\Omega_m$ and $\sigma_8$ could in principle help in breaking degeneracies between these parameters and $w$ in weak lensing observations. The CMB parameter prior probability function can be written as 
\begin{equation}
-2\log\mathcal{L}_{\rm CMB}(\bb{p}) = (\bb{p} - \bb{p}_0)^T\bb{F}_{\rm CMB}(\bb{p} - \bb{p}_0)
\end{equation}
%
where we assumed that the best fit parameters are the same $\bb{p}_0$ that appear in equation (\ref{meth:linapprox}). 
Parameter constraints from the Planck CMB experiment are made available to the public via the parameter Markov Chains (MCMC) published on the Planck Legacy Archive \footnote{The archive we used is located \url{http://pla.esac.esa.int/pla/}; we used the MCMC chains contained in the \ttt{base\_w/plikHM\_TT\_lowTEB} directory, labeled as \ttt{base\_w\_plikHM\_TT\_lowTEB\_[1-4].txt}}. We can use these MCMC data to estimate the parameter covariance matrix $\bb{\Sigma}_{\rm CMB}$ on the parameter multiplet $(\Omega_m,\Omega_bh^2,\theta_{\rm MC},\tau,w,n_s,\sigma_8)$, marginalized over the Planck nuisance parameters. We then compute the parameter prior Fisher matrix $\bb{F}_{\rm CMB} = \bb{\Sigma}_{\rm CMB}^{-1}$. Fixing the values of all parameters but $(\Omega_m,w,\sigma_8)$ and applying the prior to the weak lensing parameter likelihood (\ref{meth:paramlikelihood}) is equivalent to taking the $(\Omega_m,w,\sigma_8)$ slice of $\bb{F}_{\rm CMB}$, which we call $\bb{F}_{\rm CMB}^{(\Omega_m,w,\sigma_8)}$, and computing the parameters constraints subject to the CMB prior as

>>>>

When the dimension $N_b$ of the summary statistics space is large, numerical issues can arise in the estimation of the parameter error bars if the covariance matrix $\bbh{C}$ is measured from simulations. When $N_r$ independent realizations are used to estimate $\bbh{C}$, its inverse $\bbh{\Psi}$ is biased by a constant factor (see \citep{Hartlap07,Taylor12,Taylor14}) which can be taken into account. When the bias correction is applied, we can calculate the expectation value of the covariance estimator (\ref{meth:parcovestimator}) (see again\citep{Taylor14})
\begin{equation}
\label{meth:covdegradation}
\langle\bbh{\Sigma}\rangle = \left(\frac{N_r-N_b+N_p-1}{N_r-N_b-2}\right)\bb{\Sigma}
\end{equation}   
%
where $\bb{\Sigma}$ is the asymptotic covariance one obtains with an infinite number of realizations and $N_p=3$ is the number of parameters we are estimating. The scatter of the parameter estimates (\ref{meth:parestimate}) on the other hand scales as \citep{Taylor14}
\begin{equation}
\label{meth:estdegradation}
\langle{\rm Cov}(\bbh{p})\rangle = \left(\frac{N_r-2}{N_r-N_b+N_p-2}\right)\bb{\Sigma}
\end{equation}
%
Although equations (\ref{meth:covdegradation}) and (\ref{meth:estdegradation}) agree in the limit $N_r\rightarrow\infty$, they can be different when a finite number of realizations is used. The degradation factor in the parameter covariance estimate in (\ref{meth:covdegradation}) is of order $1+(1+N_p)/N_r$, while the scatter of the estimates $\bbh{p}$ is of order $1+(N_b-N_p)/N_r$. These numbers can be very different if $N_b$ is large. This means that the parameter error bar estimate (\ref{meth:covdegradation}) is too conservative if $N_b/N_r$ is of order unity. This could be the case with the inclusion of tomography information. If we bin the single redshift summary statistic with $N_{\rm st}$ intervals, and consider $N_z$ redshift bins, this can lead to a summary statistic vector of size $N_b = O(N_{\rm st}N_z^2)$ for the power spectrum and $N_b = O(N_{\rm st}N_z)$ for the remaining statistics. This can become quickly comparable with $N_r=16000$ once more redshift bins or a finer binning of the summary statistic are considered. In order to avoid these error degradation issues, we apply dimensionality reduction techniques to the summary statistics we are considering. Even if these techniques might not play a vital role in this work, as the maximum $N_b/N_r$ ratio we use is of order 1\%, they will definitely be relevant in future experiments when using finely binned summary statistics or when combining different cosmological probes. We explain the dimensionality reduction techniques we adopted in the next paragraph.

@@

\subsection{Dimensionality reduction}
We apply a Principal Component Analysis (see \citep{astroMLText} for example) to reduce the dimensionality of our summary statistics while preserving the cosmological information content. The model statistic $\bb{d}(\bb{p})$ can be regarded as a $P\times N_b$ matrix $d_{pi}$. Consider the whitened model matrix
\begin{equation}
\label{meth:whitening}
\Delta_{pi} = P\frac{d_{pi}}{\sum_{q=1}^Pd_{qi}} - 1.
\end{equation} 
%
We perform a Singular Value Decomposition (SVD) of $\Delta$ 
\begin{equation}
\bb\Delta = \bb{L}\bb{\Lambda} \bb{R}
\end{equation}
%
where $L$ is $P\times Q$, $\bb{\Lambda}={\rm diag}(\Lambda_1,...,\Lambda_Q)$ and $\bb{R}$ is $Q\times N_b$ and $Q={\min}(P,N_b)$. $R_{ni}$ is the $i$-th component of the $n$-th basis vector in statistics space. The singular value $\Lambda_n$ is the variance of the whitened summary statistic along the $n$-th basis vector. We assume that only summary statistic projections on the first $N_c$ basis vectors contain relevant cosmological information, where $N_c<N_b$ is a number that has to be determined from the simulations. Let $\bb{R}(N_c)$ be a matrix made of the first $N_c$ rows of $\bb{R}$ (we assume that the singular values $\Lambda_i$ are sorted from highest to lowest). We define the PCA projection of a summary statistic $\bbh{d}$ on $N_c$ principal components as 
\begin{equation}
\label{meth:pcaprojection}
\h{d}_n(N_c) = \sum_{i=1}^{N_b}(\bb{R}(N_c))_{ni}\left(P\frac{\h{d}_i}{\sum_{p=1}^P d_{pi}}-1\right) 
\end{equation} 
%
Through the above procedure, we hope to capture the cosmological information contained in $\bbh{d}$ by projecting it on the $N_c<N_b$ principal components that vary the most with cosmology parameters.

@@

\subsection{Priors from CMB experiments}
In this paragraph we describe how we included prior knowledge of cosmological parameters from previous Cosmic Microwave Background (CMB) observations, such as Planck \citep{PlanckCosmo}. CMB experiments provide tight constraints on $(\Omega_m,\sigma_8)$, but they are not sensitive to dark energy parameters such as $w$. Nevertheless, prior knowledge of $\Omega_m$ and $\sigma_8$ could in principle help in breaking degeneracies between these parameters and $w$ in weak lensing observations. The CMB parameter prior probability function can be written as 
\begin{equation}
-2\log\mathcal{L}_{\rm CMB}(\bb{p}) = (\bb{p} - \bb{p}_0)^T\bb{F}_{\rm CMB}(\bb{p} - \bb{p}_0)
\end{equation}
%
where we assumed that the best fit parameters are the same $\bb{p}_0$ that appear in equation (\ref{meth:linapprox}). 
Parameter constraints from the Planck CMB experiment are made available to the public via the parameter Markov Chains (MCMC) published on the Planck Legacy Archive \footnote{The archive we used is located \url{http://pla.esac.esa.int/pla/}; we used the MCMC chains contained in the \ttt{base\_w/plikHM\_TT\_lowTEB} directory, labeled as \ttt{base\_w\_plikHM\_TT\_lowTEB\_[1-4].txt}}. We can use these MCMC data to estimate the parameter covariance matrix $\bb{\Sigma}_{\rm CMB}$ on the parameter multiplet $(\Omega_m,\Omega_bh^2,\theta_{\rm MC},\tau,w,n_s,\sigma_8)$, marginalized over the Planck nuisance parameters. We then compute the parameter prior Fisher matrix $\bb{F}_{\rm CMB} = \bb{\Sigma}_{\rm CMB}^{-1}$. Fixing the values of all parameters but $(\Omega_m,w,\sigma_8)$ and applying the prior to the weak lensing parameter likelihood (\ref{meth:paramlikelihood}) is equivalent to taking the $(\Omega_m,w,\sigma_8)$ slice of $\bb{F}_{\rm CMB}$, which we call $\bb{F}_{\rm CMB}^{(\Omega_m,w,\sigma_8)}$, and computing the parameters constraints subject to the CMB prior as
